### 2015-06-30 Paper_v1.tex
Some initial edits

\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort,comma,numbers]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{comment}
\usepackage{lineno}
\usepackage{chicago}
\usepackage{textcomp}

@@

%\usepackage{geometry}
\usepackage{fullpage}

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large A novel metaheuristic approach to obtain near optimal solutions in high dimensional biochemical networks with a few function evaluations}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar, Christine Shoemaker and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Parameter estimation in large scale biological systems}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Biotechnology Journal}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
The problem of parameter estimation presents a significant challenge in the modeling of large complex systems. Mathematical modeling of biological systems with multiple feed back loops is one such area where parameter estimation is a difficult non-linear optimization problem. This difficulty is further compounded when dealing with parameter vectors of high dimensions. Such problems generally involve expensive function evaluations and it is hard to obtain optimal solutions within finite computational time. In this study we present a novel meta-heuristic approach that combines a variant of particle swarm optimization (PSO) along with dynamically dimensioned search (DDS) to obtain near optimal solutions of high dimensional biochemical networks within a relatively few function evaluations. We use a particle swarm optimization technique that uses multi-swarms to generate candidate vectors which are then greedily updated using DDS by dynamically varying the perturbed parameter dimensions. We tested this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other metaheuristics like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and also against DDS alone. The new algorithm outperforms all the other metaheuristics on the coagulation model. The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data. We also performed this comparison on commonly used test functions (Ackley and Rosenbrock) for global optimization and found the same behavior. Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach. We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.

@@

Biological systems interact with one another and with the external world through highly complex biochemical networks. The advent of high throughout technologies has provided researchers with a deluge of data about these networks. Over the last decade, with an increase in computational power, large scale mathematical modeling has evolved as a powerful paradigm to characterize these complex biochemical interactions \cite{assmus2006dynamics, Riel:2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}. Mathematical modeling of these networks is an iterative process that primarily consists of two tasks a) Model construction b) Model calibration or Parameter Estimation \cite{Aldridge:2006aa}\cite{banga2008optimization} \cite{ashyraliyev2009systems}. Model construction involves translating existing knowledge about the network (i.e. entities in the network, interactions amongst the entities, types of interactions) along with reasonable assumptions into an appropriate mathematical framework. Model calibration or parameter estimation follows the model construction step.  In this step the model is trained against experimental data to determine the model parameters that give a good 'fit' against experimental data. Model training is generally framed as a non-linear optimization problem wherein the residual (or objective function) between experimental data and simulation data is minimized using various optimization techniques \cite{moles2003parameter}. The optimal parameters obtained from the previous step are then used to validate the model on hitherto unseen experimental data. If validation leads to unsatisfactory results, model construction and calibration are repeated iteratively till satisfactory results are obtained.

With the capacity to construct very large models using various mathematical formalisms \cite{chen2009input} \cite{tasseff2011modeling}\cite{luan2007computationally} \cite{mo2007genome} \cite{orth2011comprehensive} \cite{karr2012whole}\cite{buchel2013path2models} \cite{smallbone2010towards}, more often than not, parameter estimation is becoming a major challenge if not a bottleneck in large scale mathematical modeling. Though parameter estimation has been a well studied problem in engineering \cite{beck1977parameter}\cite{nieman1971review}\cite{young1981parameter} \cite {beck1998inverse}, the complex dynamics of large biological systems and noisy, often incomplete experimental data pose a more difficult and rather unique challenge. Most of the optimization problems pertaining to biological systems are non-linear and multi-modal i.e. most of them have multiple local minima or maxima \cite{moles2003parameter}\cite{banga2008optimization}. The non-linearity of the problem coupled with multi-modality generally renders the local optimization techniques like pattern search method \cite{hooke1961direct}, Nelder-Mead simplex method \cite{nelder1965simplex} or Levenberg-Marquardt method\cite{more1978levenberg} incapable of obtaining an optimal solution. Though deterministic global optimization techniques can handle non-linearity and multi-modality \cite{esposito2000deterministic} \cite{horst2013global}, the absence of derivative information, discontinuity of the objective functions, non-smooth regions or the lack of any knowledge about the objective function severely hamper the use of these techniques.

Metaheuristic approaches like Genetic Algorithms, Simulated Annealing \cite{kirkpatrick1983optimization}, Evolutionary Programming and population based searches like Differential Evolution\cite{storn1997differential} have shown promise in this regard \cite{sun2012parameter}. They do not make any assumptions about the structure of objective function or require any \textit{a priori} information about the objective function. Though they do not guarantee strong convergence, these approaches are effective in finding near optimal solutions.  Mendes et al. \cite{mendes1998non} used Simulated Annealing to estimate rate constants for the irreversible inhibition of HIV proteinase, Modchang et al. \cite {modchang2008mathematical} used Genetic Algorithms to estimate parameters for a mathematical model of signal transduction, Differential Evolution based approaches have been an effective approach on various systems \cite{tsai2005evolutionary}, \cite{wang2001hybrid},\cite{noman2007inferring}. Tashkova et al. \cite{tashkova2011parameter} compared different metaheuristics for parameter estimation on a dynamic model of endocytosis and showed that DE was the most effective. Banga and co-workers have applied scatter-search based methods \cite{villaverde2012cooperative} \cite{rodriguez2006novel} \cite{egea2007scatter} to estimate parameters on non-linear biological processes. Hybrid approaches that combine a metaheuristic with a local optimization search, wherein a near optimal solution that is obtained with a metaheuristic is further refined using a local search have also become popular. Villaverde et al. \cite{villaverde2015biopredyn} combined scatter search with local search methods for parameter estimation in large scale systems biology models. Fan et al. recently showed that population based metaheuristics along with decomposition based methods can be used to model gene circuits from mRNA data \cite{fan2015parameter}.

Despite these successes, a  major drawback with most of these approaches is the vast number of objective function evaluations they take in arriving at good search regions. As the models grow in size and complexity and the number of dimensions of parameter vector increases,  parameter estimation becomes a computationally expensive task. In many of these high dimensional problems approaching an exact solution may not be necessary. Gutenkust et al. \cite{gutenkunst2007universally} showed that a number of systems biology models are 'sloppy'. Sloppy systems have specific parameter combinations that largely define the dynamics of the system. Large perturbations to the rest of the parameters does not greatly impact the system dynamics. Ensemble approaches \cite{song2010ensembles,luan2010ensembles} have exploited this aspect to describe the dynamics of biological systems including coagulation which can be described using only a set of key species or parameters \cite{sagar2015dynamic}. Tolson and Shoemaker \cite{tolson2007dynamically} showed through Dynamically Dimensioned Search (DDS) that high-dimensional watershed models can be calibrated quickly by perturbing only a subset of dimensions. Thus in our current study we took into cognizance the fact that it may not be necessary to obtain the exact solution for high dimensional biological systems and that good enough solutions can be quickly obtained without expending a lot of objective function evaluations. Hence our current approach uses the power of population based heuristics along with DDS to obtain near optimal or good solutions. Though in theory we can combine any metaheuristic with DDS, for the current purpose we used Particle Swarm Optimization (PSO). PSO unlike Genetic Algorithm (GA) or Differential Evolution (DE) does not have complex operations like cross over, mutation or recombination. It is simple to use and does not have a lot of parameters associated with other heuristics. However PSO is known to rapidly converge to a local optimum and thus several variants \cite{peer2003using,zhan2009adaptive,li2007fast} have been developed of which Multi swarm particle swarm optimization approaches (MLSPSO) \cite{zhao2008dynamic,liang2005dynamic} are one of the effective ones. We used MLSPSO to preclude bad regions of search and thereafter search within these regions using DDS.

@@

\section*{Results}
\subsection*{Performance on Coagulation model}
Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops (Figure 1). The biochemistry underlying coagulation, though quite complex has been well studied \cite{mann2003dynamics,mann2003all,mann2003thrombin,vogler2009contact,diamond2013systems,fogelson2005coagulation,anand2003model}, and reliable experimental coagulation models have been developed \cite{hockin2002model,chatterjee2010systems,mann2006models,luan2007computationally}. This makes it an ideal system for mathematical modeling and parameter estimation. Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets. The coagulation factors are generally in an inactive state and are known as zymogens. These zymogens are activated through certain triggers. Trigger events like injury or trauma or sepsis expose factors like collagen, tissue factor and von Willebrand factor (vWF) to blood. The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.   Luan et al. modeled coagulation using coupled non-linear ordinary differential equations with 148 reactions and 92 species \cite{luan2007computationally} and validated model using 21 published datasets.

We compared the performance of Dynamically Dimensioned Search with Multi Swarm Particle Swarm Optimization (DDSMLSPSO) on this model of coagulation against commonly used meta heuristics like Simulated Annealing (SA), Differential Evolution (DE), Particle Swarm Optimization (PSO) and also against Dynamically Dimensioned Search (DDS). To train the model parameters, we used data sets from TF-VIIa initiated coagulation with no anticoagulants. The objective function is a weighted linear combination of two different error functions that used data sets representing coagulation initiated with different concentrations of TF-VIIa (5pM, 5nM) \cite{hockin2002model}. We restricted the number of function evaluations to 4000 for each algorithm and performed 25 trials of this experiment. DDSMLSPSO exhibits a much faster rate of error convergence and has a much lower final error as compared to the other algorithms (Figure 3). Within the first 1000 function evaluations (swarm phase) of DDSMLSPSO there is a very rapid drop in error. Subsequently, again after 2000 function evaluations (dynamically dimensioned phase) the error drops quickly. Overall at the end of 4000 function evaluations DDSMLSPSO minimizes the error (final objective error is 90456.9) to a much greater extent than any of the other algorithms. Amongst the rest of the algorithms DDS has a comparable error drop for the first 2000 function evaluations. However the drop in error remains constant for DDS after 2000 function evaluations. Using the parameters obtained at the end of 4000 function evaluations we examined the 'fits' between models predictions and experimental data (Figure 2). The solid lines represent the mean value of prediction over 25 trials and the shaded region represents the 99\% confidence interval. Subsequently we compared the model predictions against completely 'unseen' or untrained experimental where coagulation was initiated with 500pM,50pM,10pM concentrations of TF-VIIa respectively (Figure 3).

\subsection*{Performance on Benchmark functions}

We compared the performance of these algorithms on commonly used test functions for global optimization. We used a 300 dimensional Rastrigin function and 300 dimensional Ackley function. Both Ackley and Rastrigin have multiple local minima and maxima and attain a global minimum value of 0. We tested the performance of the 5 different heuristic approaches on these two functions. In each experiment we again fixed the number of function evaluations at 4000 and ran 25 experiments. In both cases (Figure S6 A and B) we see that the error convergence rate for DDSMLSPSO is much faster as compared to the other four metaheuristics and it finds the global minimum of 0 in both cases.

@@

\begin{table}[h]
\centering
\caption{Table with optimization settings and results for coagulation and different benchmarks using DDSMLSPSO.}
\label{table:1}
\resizebox{\columnwidth}{!}{
\begin{tabular}{|cccccc|}
\hline
                        & Coagulation           & \multicolumn{4}{l}{Benchmarks}\\
 \hline  
                        & Experimental Data Set & B1           & B4          & Ackley 300D & Rastrigin 300D \\
 \hline                        
Lower Bound             & 0.001.pnom            & 5.pnom       & 5.pnom      & 30          & 5.12           \\
Upper Bound             & 1000.pnom             & 0.2.pnom     & 0.2.pnom    & -15         & -5.12          \\
CPU Time                & 10.0835 hours         & 38.308 hours & 6.2 minutes & 2.8 seconds & 2.6 seconds    \\
Function Evaluations    & 4000                  & 4000         & 4000        & 4000        & 4000           \\
Initial Objective Value & 1.11837e+07           & 1.4275e+07   & 1.8536e+07  & 21.12       & 99.985         \\
Final Objective Value   & 90456.9               & 4.96e+05     & 38.9375     & 0           & 0              \\
Nominal Objective Value & 4.7785e+06            & 1.0986e+06   & 39.0676     & 0           & 0      \\       
\hline
\end{tabular}
}
\end{table}

\clearpage
\section*{Discussion}
Our study presents a novel approach for high-dimensional parameter estimation in complex biological systems with relatively few function evaluations. In this approach we combined a variant of a well known meta heuristic particle swarm optimization with Dynamically Dimensioned Search (DDS). We tested our approach on an ODE model of coagulation with 148 parameters and 92 species. Coagulation is an ideal system to test our approach since the biology is well known and complex, with multiple feed back loops that are tightly regulated. We used experimental data with under different conditions to obtain optimal parameters and used these parameters to make predictions against unseen experimental data. We obtained good fits and made accurate enough predictions using parameters obtained from 4000 function evaluations. Further, we also used high-dimensional forms of commonly used test functions of global optimization and showed that we were able to find the global minimum for 300 dimensional Ackley and Rastrigin functions faster than other metaheuristics. We also considered two recently published benchmark problems to test parameter optimization approaches and showed that we were able to retrieve the nominal parameter vector within 4000 function evaluations.

>>>>

\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses
\usepackage{ifthen}  % Conditional
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort,comma,numbers,sort&compress]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{fullpage}
\usepackage{comment}
\usepackage{lineno}
%\usepackage{chicago}
\usepackage{textcomp}

@@

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Dynamically dimensioned particle swarm optimization (DdPSO): A meta-heuristic for parameter estimation in large biochemical models.}}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar, Christine Shoemaker and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Parameter estimation in biochemical models}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Biotechnology Journal}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
The problem of parameter estimation presents a significant challenge in the modeling of large complex systems. Mathematical modeling of biological systems with multiple feed back loops is one such area where parameter estimation is a difficult non-linear optimization problem. This difficulty is further compounded when dealing with parameter vectors of high dimensions. Such problems generally involve expensive function evaluations and it is hard to obtain optimal solutions within finite computational time. In this study we present a novel meta-heuristic approach that combines a variant of particle swarm optimization (PSO) along with dynamically dimensioned search (DDS) to obtain near optimal solutions of high dimensional biochemical networks within a relatively few function evaluations. We use a particle swarm optimization technique that uses multi-swarms to generate candidate vectors which are then greedily updated using DDS by dynamically varying the perturbed parameter dimensions. We tested this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and also against DDS alone. The new algorithm outperforms all the other meta-heuristics on the coagulation model. The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data. We also performed this comparison on commonly used test functions (Ackley and Rosenbrock) for global optimization and found the same behavior. Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach. We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.

@@

Biological systems interact with one another and with the external world through highly complex biochemical networks. The advent of high throughout technologies has provided researchers with a deluge of data about these networks. Over the last decade, with an increase in computational power, large scale mathematical modeling has evolved as a powerful paradigm to characterize these complex biochemical interactions \cite{assmus2006dynamics, Riel:2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}. Mathematical modeling of these networks is an iterative process that primarily consists of two tasks a) Model construction b) Model calibration or Parameter Estimation \cite{Aldridge:2006aa,banga2008optimization,ashyraliyev2009systems}. Model construction involves translating existing knowledge about the network (i.e. entities in the network, interactions amongst the entities, types of interactions) along with reasonable assumptions into an appropriate mathematical framework. Model calibration or parameter estimation follows the model construction step.  In this step the model is trained against experimental data to determine the model parameters that give a good 'fit' against experimental data. Model training is generally framed as a non-linear optimization problem wherein the residual (or objective function) between experimental data and simulation data is minimized using various optimization techniques \cite{moles2003parameter}. The optimal parameters obtained from the previous step are then used to validate the model on hitherto unseen experimental data. If validation leads to unsatisfactory results, model construction and calibration are repeated iteratively till satisfactory results are obtained.

With the capacity to construct very large models using various mathematical formalisms \cite{chen2009input,tasseff2011modeling,luan2007computationally,mo2007genome,orth2011comprehensive,karr2012whole,buchel2013path2models,smallbone2010towards}, more often than not, parameter estimation is becoming a major challenge if not a bottleneck in large scale mathematical modeling. Though parameter estimation has been a well studied problem in engineering \cite{nieman1971review,beck1977parameter,young1981parameter,beck1998inverse}, the complex dynamics of large biological systems and noisy, often incomplete experimental data pose a more difficult and rather unique challenge. Most of the optimization problems pertaining to biological systems are non-linear and multi-modal i.e. most of them have multiple local minima or maxima \cite{moles2003parameter,banga2008optimization}. The non-linearity of the problem coupled with multi-modality generally renders the local optimization techniques like pattern search method \cite{hooke1961direct}, Nelder-Mead simplex method \cite{nelder1965simplex} or Levenberg-Marquardt method\cite{more1978levenberg} incapable of obtaining an optimal solution. Though deterministic global optimization techniques can handle non-linearity and multi-modality \cite{esposito2000deterministic,horst2013global}, the absence of derivative information, discontinuity of the objective functions, non-smooth regions or the lack of any knowledge about the objective function severely hamper the use of these techniques.

Meta-heuristic approaches like Genetic Algorithms, Simulated Annealing \cite{kirkpatrick1983optimization}, Evolutionary Programming and population based searches like Differential Evolution\cite{storn1997differential} have shown promise in this regard \cite{sun2012parameter}. They do not make any assumptions about the structure of objective function or require any \textit{a priori} information about the objective function. Though they do not guarantee strong convergence, these approaches are effective in finding near optimal solutions.  Mendes et al. \cite{mendes1998non} used Simulated Annealing to estimate rate constants for the irreversible inhibition of HIV proteinase, Modchang et al. \cite {modchang2008mathematical} used Genetic Algorithms to estimate parameters for a mathematical model of signal transduction, Differential Evolution based approaches have been an effective approach on various systems \cite{tsai2005evolutionary,wang2001hybrid,noman2007inferring}. Tashkova et al. \cite{tashkova2011parameter} compared different meta-heuristics for parameter estimation on a dynamic model of endocytosis and showed that DE was the most effective. Banga and co-workers have applied scatter-search based methods \cite{villaverde2012cooperative,rodriguez2006novel,egea2007scatter} to estimate parameters on non-linear biological processes. Hybrid approaches that combine a meta-heuristic with a local optimization search, wherein a near optimal solution that is obtained with a meta-heuristic is further refined using a local search have also become popular. Villaverde et al. \cite{villaverde2015biopredyn} combined scatter search with local search methods for parameter estimation in large scale systems biology models. Fan et al. recently showed that population based meta-heuristics along with decomposition based methods can be used to model gene circuits from mRNA data \cite{fan2015parameter}.

Despite these successes, a  major drawback with most of these approaches is the vast number of objective function evaluations they take in arriving at good search regions. As the models grow in size and complexity and the number of dimensions of parameter vector increases,  parameter estimation becomes a computationally expensive task. In many of these high dimensional problems approaching an exact solution may not be necessary. Gutenkust et al. \cite{gutenkunst2007universally} showed that a number of systems biology models are 'sloppy'. Sloppy systems have specific parameter combinations that largely define the dynamics of the system. Large perturbations to the rest of the parameters does not greatly impact the system dynamics. Ensemble approaches \cite{song2010ensembles,luan2010ensembles} have exploited this aspect to describe the dynamics of biological systems including coagulation which can be described using only a set of key species or parameters \cite{sagar2015dynamic}. Tolson and Shoemaker \cite{tolson2007dynamically} showed through Dynamically Dimensioned Search (DDS) that high-dimensional watershed models can be calibrated quickly by perturbing only a subset of dimensions. Thus in our current study we took into cognizance the fact that it may not be necessary to obtain the exact solution for high dimensional biological systems and that good enough solutions can be quickly obtained without expending a lot of objective function evaluations. Hence our current approach uses the power of population based heuristics along with DDS to obtain near optimal or good solutions. Though in theory we can combine any meta-heuristic with DDS, for the current purpose we used Particle Swarm Optimization (PSO). PSO unlike Genetic Algorithm (GA) or Differential Evolution (DE) does not have complex operations like cross over, mutation or recombination. It is simple to use and does not have a lot of parameters associated with other heuristics. However PSO is known to rapidly converge to a local optimum and thus several variants \cite{peer2003using,zhan2009adaptive,li2007fast} have been developed of which Multi swarm particle swarm optimization approaches (MLSPSO) \cite{zhao2008dynamic,liang2005dynamic} are one of the effective ones. We used MLSPSO to preclude bad regions of search and thereafter search within these regions using DDS.

@@

\section*{Results}
\subsection*{Dynamically dimensioned particle swarm optimization (DdPSO).}
DdPSO is a novel meta-heuristic that combines multi-swarm based particle swarm methods with dynamically dimensioned search (Fig. \ref{fig-algorithm}).
The goal of this approach is to obtain optimal or near optimal parameters for high-dimensional complex biological systems within a pre-specified number of function evaluations. We randomly initialized a swarm of $\mathcal{K}$-dimensional particles (represented as ${x}_{i}$), wherein each of these particles corresponds to a $\mathcal{K}$-dimensional parameter vector. After initialization, the particles were grouped into different sub-swarms randomly. Thereafter within each sub-swarm ${S}_{k}$,  particles were updated according to the following rule.
\begin{eqnarray}
	\mathbf{x}_{i,j} &=&\theta_{1,j-1}\mathbf{x}_{i,j-1} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{x}_{i,j-1}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{GL}_{k} - \mathbf{x}_{i,j-1}\right)
\end{eqnarray}
where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the best local solution found by particle $i$ till function evaluation $j$-$1$, and
$\mathcal{GL}_{k}$ denotes the best local solution found over the population of all particles within the swarm ${S}_{k}$. The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
parameters ($\mathcal{K}\times{1}$). In our algorithm the parameter $\theta_{1,j-1}$ depends on the function evaluations and is controlled according to the following equation
\begin{eqnarray}
	\mathbf \theta_{1,j}&=&((\mathbf{N}-{j})*(\mathbf{w}_{max}-\mathbf{w}_{min}))/(\mathbf{N}-{1}) + \mathbf{w}_{min}
\end{eqnarray}
where $\mathbf{N}$ represents the total number of function evaluations, $\mathbf{w}_{max}$ and $\mathbf{w}_{min}$ are the maximum and minimum inertia weights respectively. While updating the particle, we made sure all dimensions of the solution represented by the particle were within bounds using a reflection boundary condition. After every $\mathbf{g}$  function evaluations, the particles within all sub-swarms were mixed and then randomly redistributed to a new sub-swarm. The particles were then again updated according Eq.1. This process continued till $\mathcal{FR}*\mathbf{N}$ number of functions evaluations, where $\mathcal{FR}$ represents the fraction of evaluations with the multi-swarms. At the end of these function evaluations, we froze all the solutions represented by various particles and chose the particle with best solution among $\mathcal{GL}_{1} \cdots \mathcal{GL}_{NS}$ as the initial candidate vector $\mathcal{G}$ for the remaining $({1}-\mathcal{FR})*\mathbf{N}$ number of function evaluations.

This particle was then updated according to the following rule
\begin{eqnarray}
	\mathcal{G}_{new}(\mathbf{J})& = &\mathcal{G}(\mathbf{J})+\mathbf{r}_{normal}(\mathbf{J})\sigma(\mathbf{J})
\end{eqnarray}
where $\mathbf{J}$ represents the set containing the specific dimensions being perturbed, ${r}_{normal}$ denotes a normal random vector of the same dimensions as $\mathcal{G}$. $\sigma$ is the amplitude of perturbation given by following equation:
\begin{eqnarray}
	\sigma& = &\mathbf{R}(\mathcal{MAX} -\mathcal{MIN})
\end{eqnarray}
where $\mathbf{R}$ is the scalar perturbation size parameter, $\mathcal{MAX}$ and $\mathcal{MIN}$ are ($\mathcal{K}\times{1}$) vectors that represent the maximum and minimum bounds on each dimension. The set $\mathbf{J}$ was constructed using a probability function $\mathcal{P}_{j}$ that represented the probability whether a specific dimension $j$ was perturbed or not.  This function is a monotonically decreasing function that decreases with the number of function evaluations. $\mathcal{P}_{j}$ can be any monotonically decreasing function, in our approach we used the following function:
\begin{eqnarray}
	\mathcal{P}_{j}&=&{1}-\log(j/ (\beta({1}-\mathcal{FR})*\mathbf{N}))
\end{eqnarray}
where $\beta$ is the perturbation frequency probability modulator. Thus the number of dimensions of the candidate vector that are updated or perturbed decreases with the as the number of function evaluations increase. These updates are greedy in nature, so $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than the old one $\mathcal{G}$.

\subsection*{Performance of DdPSO on benchmark problems.}
We compared the performance of DdPSO with algorithms on commonly used test functions for global optimization, as well as benchmark biochemical problems from literature (Table \ref{table:benchmark-problems}). We used a 300 dimensional Rastrigin function and 300 dimensional Ackley function. Both Ackley and Rastrigin have multiple local minima and maxima and attain a global minimum value of 0. We tested the performance of the 5 different heuristic approaches on these two functions. In each experiment we again fixed the number of function evaluations at 4000 and ran 25 experiments. In both cases (Figure S6 A and B) we see that the error convergence rate for DDSMLSPSO is much faster as compared to the other four s and it finds the global minimum of 0 in both cases.

@@

\subsection*{Performance of DdPSO on a model of the human coagulation cascade.}
We compared the performance of DdPSO on a model of blood coagulation against commonly used meta-heuristics like simulated annealing (SA), differential evolution (DE), particle swarm optimization (PSO),and dynamically dimensioned search (DDS). Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops (Fig. \ref{fig-coagulation-network}). The biochemistry underlying coagulation, though quite complex has been well studied \cite{mann2003dynamics,mann2003all,mann2003thrombin,vogler2009contact,diamond2013systems,fogelson2005coagulation,anand2003model}, and reliable experimental coagulation models have been developed \cite{hockin2002model,chatterjee2010systems,mann2006models,luan2007computationally}. This makes it an ideal system for mathematical modeling and parameter estimation. Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets. The coagulation factors are generally in an inactive state and are known as zymogens. These zymogens are activated through certain triggers. Trigger events like injury or trauma or sepsis expose factors like collagen, tissue factor and von Willebrand factor (vWF) to blood. The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.   Luan et al. modeled coagulation using coupled non-linear ordinary differential equations with 148 reactions and 92 species \cite{luan2007computationally} and validated model using 21 published datasets.

 To train the model parameters, we used data sets from TF-VIIa initiated coagulation with no anticoagulants. The objective function was a weighted linear combination of two different error functions that used data sets representing coagulation initiated with different concentrations of TF-VIIa (5pM, 5nM) \cite{hockin2002model}. We restricted the number of function evaluations to 4000 for each algorithm and performed 25 trials of this experiment. DdPSO exhibits a much faster rate of error convergence and has a much lower final error as compared to the other algorithms (Figure 3). Within the first 1000 function evaluations (swarm phase) of DDSMLSPSO there is a very rapid drop in error. Subsequently, again after 2000 function evaluations (dynamically dimensioned phase) the error drops quickly. Overall at the end of 4000 function evaluations DDSMLSPSO minimizes the error (final objective error is 90456.9) to a much greater extent than any of the other algorithms. Amongst the rest of the algorithms DDS has a comparable error drop for the first 2000 function evaluations. However the drop in error remains constant for DDS after 2000 function evaluations. Using the parameters obtained at the end of 4000 function evaluations we examined the 'fits' between models predictions and experimental data (Figure 2). The solid lines represent the mean value of prediction over 25 trials and the shaded region represents the 99\% confidence interval. Subsequently we compared the model predictions against completely 'unseen' or untrained experimental where coagulation was initiated with 500pM,50pM,10pM concentrations of TF-VIIa respectively (Figure 3).

\clearpage
\section*{Discussion}
Our study presents a novel approach for high-dimensional parameter estimation in complex biological systems with relatively few function evaluations. In this approach we combined a variant of a well known meta heuristic particle swarm optimization with Dynamically Dimensioned Search (DDS). We tested our approach on an ODE model of coagulation with 148 parameters and 92 species. Coagulation is an ideal system to test our approach since the biology is well known and complex, with multiple feed back loops that are tightly regulated. We used experimental data with under different conditions to obtain optimal parameters and used these parameters to make predictions against unseen experimental data. We obtained good fits and made accurate enough predictions using parameters obtained from 4000 function evaluations. Further, we also used high-dimensional forms of commonly used test functions of global optimization and showed that we were able to find the global minimum for 300 dimensional Ackley and Rastrigin functions faster than other meta-heuristics. We also considered two recently published benchmark problems to test parameter optimization approaches and showed that we were able to retrieve the nominal parameter vector within 4000 function evaluations.
